---
title: "session09 - exercises"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

```

# Introduction

The main **learning goals** of this week's practical parts are:

-   getting an intuition about the gap between verbal hypotheses and formal hypotheses

# The data

In this dataset, we will be analyzing data from @kilpatrick2023random:

In a nutshell, the authors were interested in whether there is a relationship between the form of a word and its meaning by looking at Japanese Pokemon names.

Let's have a look at the data:

```{r}
#| label: data

# Load in the dataset and already remove irrelevant stuff:
pokemon <- read_csv('../data/Japanese_Pokemon_simplified.csv') |> 
  select(-`Sum of values`, -Checksum)

# check data set:
pokemon

```

What you see here are the following variables:

Variables:

`Designation`: source of Pokemon name. Official = the official names of Japanese Pokemon; Elicited = elicited names from naive participants.

(`pokemonname`: Pokemon name in Japanese) removed in simplified data file

`Output`: Pokemon name in transcribed form

`Length`: Amount of phones in name

`Evolution`: Evolution stage of Pokemons: 1 = pre-evolution; 2 = post-evolution, NA = do not evolve

`f` = amount of /f/ sounds in name

`T` = amount of /T/ sounds in name

...

and many more different sounds.

Now imagine researchers want to use this data set to *test* the following hypothesis:

*The evolution stage of a Pokemon is reflected in aspects of its Japanese name*

# Exercise

## A: Preregister your analysis

Take the hypothesis and what you know about the data set and its structure to predict a very specific relationship that you can test. You are allowed to look at the dataset, but do not look at concrete relationships that might be a manifestation of the hypothesis.

Is the length of a Japanese Pokemon name related to its evolution stage?

HYPOTHESIS: Pokemon with longer names are more likely to be evolution 2 than evolution 1 or a pokemon that do not evolve ( aka NA in the Evolution category)

Pokemon in evolution stage 2 have longer names than Pokemon in stage 1 or that do not evolve.

I made the hypothesis that Pokemon with longer names are more likely to be either in Evolution stage 2 or in the non-evolving category than in evolution stage 1. To test this, I first subset the data into these important pieces I am using. Then, I coded the subset with Evolution 2 being "1" , anything that is not 2 , aka Evolution 1 or NA, being "0". Then I ran a model on this new little "Stage 2" binomial factor and see what comes from it.

## B: Run your preregistered analysis

Run the preregistered analysis made in (A) above and interpret your results relative to the hypothesis.

```{r}

# MAKING SUBSET
pokemon_subset <- subset(pokemon, Evolution %in% c("1", "2", "NA"))

# BINOMIAL MAKING FOR THE MODEL
pokemon_subset$Stage2 <- ifelse(pokemon_subset$Evolution == "2", 1, 0)

# MODEL WITH BINOMIAL AND SUBSET 
model_1 <- glm(Stage2 ~ Length, data = pokemon_subset, family = binomial())

# LET'S TAKE A PEEK AT THIS MODEL SHALL WE
summary(model_1)
```

#### The results for the Length coefficient are :

estimate - 0.36455

standard error - 0.03075

z value - 11.85

p value - \< 2e-16

### What this means for the model -

-   The effect is positive!!!! longer names increase the probability of being stage 2.... but is it significant

-   YES. The effect is highly statistically significant look at the P value ????

This means that onec can determine longer Japanese Pokemon names can be associated with a higher probability of being in the Evolution 2 category rather than Evolution 1 or a non-evolving Pokemon!

But...... let's challenge this

## C: Show contradicting results

Now explore the data and find two alternative way to test the research hypothesis that yield convincing results that go in opposite directions based of your findings in (A). Justify all your choices verbally after the fact.

```{r}

# Create vowels in the SAME dataset you're modeling
newish_subset <- pokemon_subset %>% 
  mutate(vowels = a + e + i + o + u + R)

# Run model
model_c2 <- glm(Stage2 ~ vowels + Length, 
                data = newish_subset, 
                family = binomial())

summary(model_c2)
exp(coef(model_c2))

```

#mo vowels mo evolution .......

I wanted to look if there was anything that could be affecting Evolution, not just length alone, but something underneath truly driving this "length" factor. This is claiming maybe it is not length that determines level of Evolution, but possibly vowel number instead.

```{r}


model_c3 <- glm(Stage2 ~ f + T + b + d + g + G + z + j + p + t + C + k + K + s +S + h + m + M + n + N + r + L + y + w + Q + X + H + v + P + c, 
                data = newish_subset_2, 
                family = binomial())

summary(model_c3)
exp(coef(model_c3))
```

After running this model, although it doesn't prove my previous hypothesis wrong, it shows a large variate of P values based off of each predictor. Now, if I knew more about the Japanese language, maybe I could specify what type of consonants/vowels/etc that could appear as an effect. However, at my sort of "blind" state, I am not sure how to interpret this fully.

## D: Researcher degrees of freedom

Note down all dimensions of researcher degrees of freedom that make the mapping of the hypothesis onto your data analysis vague. How many possible combinations can you think of?

I think that the wording of the goal of the authors is quite vague, which I believe leaves the hypothesis very open for interpretation. I would guesstimate there could be hundreds of ways to interpret this based off of model possibilities, random slopes and intercepts, etc. - all of these predictors can change so much of the mapping of this hypothesis.

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
# 
:::
